---
title: "demo"
author: "Wojciech Zajkowski"
date: "2026-03-15"
output: html_document
---

# Demo: How to Run the Reproducibility Workflow 

This demo provides minimal steps to run key parts of the codebase with included data. Full reproducibility pipeline can be found in the `replicate_part1.Rmd` and `replicate_part2.Rmd`files.

## Inputs used in the demo

- Behavioral datasets from `0.data/` loaded via helper functions.
- Precomputed model outputs from `2.models/fits/loo/` and `2.models/fits/preds/`.

## Demo A: Behavioral replication checks

Run in R from repository root:

```{r}
library(tidyverse)
library(cibtools)
library(wztools)
source("1.scripts/helper_funcs.R")

all <- getDataR(type = "main", path = "0.data")
rall <- getDataR(type = "rat", path = "0.data")

# Example: size-rating vs objective-size consistency for Experiment 5
size_corrs <- getSizeAccR(rall[[5]], exp = 5, method = "spearman", path = "0.data")
summary(size_corrs$size_est)
```


### Typical runtime

- Data loading + one summary check: **<1 minute** on a current computer.

## Demo B: Model output replication checks (without refitting)

Run in R:

```{r}
library(tidyverse)

loo_e2 <- read_csv("2.models/fits/loo/loo_table_E2.csv")
preds_e2 <- read_csv("2.models/fits/preds/qual_preds_E2.csv")

print(loo_e2)
print(head(preds_e2))
```


### Typical runtime

- Loading precomputed `.csv` model outputs: **<1 minute**.

## Optional: Full notebook execution

- `replicate_part1.Rmd` (behavioral analyses): typically minutes to tens of minutes depending on environment.
- `replicate_part2.Rmd` (model-based analyses using saved outputs): typically minutes.
- Full model **refitting** is computationally heavier and may take substantially longer (hours) depending on hardware and settings.

## Reproducibility expectation

Running the demo commands should reproduce key summary structures and loaded model output tables using the included datasets.
