---
title: "Replicating CIB Results"
subtitle: "Part 2: Generative Modelling"
author: Wojciech Zajkowski
date: "2026-03-01"
output: html_document
---

This notebook covers replication of the model-based analysis. Stan code for all fitted models is provided in this repository in the *2.models/stan_code*. 

## Setup

```{r}
library(cibtools)
library(wztools)
library(tidyverse)

fitted_mods_path = '2.models/fits'

```
for plotting:
```{r}
setGGoptions()
red1  ='#D54646'
red2  ='#F8766D'
red3 = '#fc9272'
blue1 ='#235FB9'
blue2 ='#00BFC4'
blue3 = 'lightblue'
ctrue = '#1b9e77' #for data
cfull= '#e7298a' #for full model predictions
ccb = '#d95f02' #for CB-only model predictions
cvu = '#7570b3' #for VU-only model predictions
main_cols=c('#D54646','#F8766D','#235FB9','#00BFC4')
```

##  Model fit scores

Because the draw files are too large to store in an online repository (~.5-2GB per model), here we will load lists of already precalculated LOO scores.
```{r}
exp_num=3
file_path = paste0(fitted_mods_path,'/loo/loo_table_E',exp_num, '.csv')
loo = read_csv(file_path)

```
Each element of the list corresponds to a single model, e.g., full model:
```{r}
loo
```

## Qualitative Predictions

Load preditions file
```{r}
exp_num=2
file_path = paste0(fitted_mods_path,'/preds/qual_preds_E',exp_num, '.csv')
x = read_csv(file_path)
x
```
This file contains model-derived predictions, predicting choices (dpred) and judgments (jpred)  for 3 model types: full, CB-only (cb) and VU-only (vu)

xx it would be nice to filter out unnecessary cols...

### Choice-Induced Bias

Comparing CIB from data (real_cb) to full model prediction (pred_cb)
```{r}
cb_tab = x %>% filter(choice>-1) %>% group_by(subjID, dec_type, judge_type) %>%
  summarise(true_cb = mean(cbias, na.rm=T),
            pred_cb = mean(cbias_pred, na.rm=T))
cb_tab %>% 
  group_by(dec_type, judge_type) %>% 
  summarise(real_cb = mean(true_cb),
            pred_cb = mean(pred_cb))
```

```{r}
#prepare for plotting
cb_tab_long = cb_tab %>% 
  filter(dec_type!=2) %>%  #remove fcd (for now)
  pivot_longer(cols = c(4:5), 
               values_to = 'cons_value',
               names_to = 'cons_type') 
cb_tab_long$true_bin=ifelse(cb_tab_long$cons_type=='true' ,T, F) 
cb_tab_long$con = cb_tab_long$dec_type==cb_tab_long$judge_type
cb_tab_long$cons_type = as.factor(cb_tab_long$cons_type)
#add this weird interaction variable for plotting (judge x con_type)
 cb_tab_long$jxc = interaction(cb_tab_long$judge_type,cb_tab_long$cons_type)
 a = levels(cb_tab_long$jxc) #original levels
 cb_tab_long$jxc = factor(cb_tab_long$jxc ,
                             levels= c(a[3:4] ,a[1:2]))
 cb_tab_long$con = factor(cb_tab_long$con, levels = c(T, F))
 cb_tab_long$cxd = interaction(cb_tab_long$con, cb_tab_long$cons_type)
orig_levels = levels(cb_tab_long$cxd)
cb_tab_long$cxd  = factor(cb_tab_long$cxd ,
                             levels = c(orig_levels[3:4], orig_levels[1:2]))
#and again...
cb_tab_long$cxj = interaction(cb_tab_long$con, cb_tab_long$judge_type)
orig_levels2 = levels(cb_tab_long$cxj)
cb_tab_long$cxj  = factor(cb_tab_long$cxj ,
                             levels = c(orig_levels2[3:4], orig_levels2[1:2]))
```

```{r}
pos <- position_dodge(0.9)
 cb_tab_long  %>% 
  filter(dec_type!='fcd') %>% 
  #filter(!is.na(dec_type)) %>%
  ggplot(aes(x=rev(as.factor(con)), y=cons_value, group=cxd, color=cxj,  linetype=cons_type), position=pos)+
   geom_hline(yintercept = 0, linetype='dashed', linewidth=2) +
   geom_violin(aes(fill=cxj, alpha=cons_type) ,position=pos) +
   geom_boxplot(aes(size=cons_type),position=pos) +
  facet_wrap(~judge_type)+
  theme_classic() +
 # facet_wrap(~con) +
#  scale_size_manual(values=c(1,2))+
  scale_size_manual(values=c(.75,1.25))+
   scale_alpha_manual(values=c(.6,1))+
  scale_linetype_manual(values=c('dotted', 'solid')) +

  scale_fill_manual(values=c(blue1, blue2,red1, red2)) +
 scale_color_manual(values=c(blue1, blue2,red1, red2))+
  ylab(label = '') +
  xlab(label = '') +
  theme_classic(base_size = 46)+
  theme(
    strip.background = element_blank(),
    legend.position = 'none',
    axis.text.x = element_blank(),
    strip.text = element_blank()
  ) 
```

### Choice consistencies

```{r}
cons_tab = x %>% filter(choice>-1) %>% group_by(subjID, dec_type, judge_type) %>%
  summarise(true_acc = mean(acc, na.rm=T),
            pred_full_acc = mean(pred_full_acc, na.rm=T),
            pred_cb_acc = mean(pred_cb_acc, na.rm=T),
            pred_vu_acc = mean(pred_vu_acc, na.rm=T)) 
cons_tab %>% 
  filter(dec_type!='fcd') %>% 
  group_by(dec_type, judge_type) %>% 
  summarise(real_cons = mean(true_acc),
            pred_cons_full = mean(pred_full_acc),
            pred_cons_cb  = mean(pred_cb_acc),
            pred_cons_vu  = mean(pred_vu_acc))
```
We can then compare the full model predictions to the actual data:
```{r}
cons_tab_long = 
  cons_tab %>% 
  filter(!is.na(dec_type)) %>% 
  pivot_longer(cols = 4:7, names_to='pred_type', values_to = 'value') %>% 
  mutate(
    pred_type = as_factor(pred_type),
  )
#binary: true or model pred
cons_tab_long$true_bin=ifelse(cons_tab_long$pred_type=='true_acc' ,T, F) 
cons_tab_long$con = as_factor(ifelse(cons_tab_long$dec_type==cons_tab_long$judge_type, 'congruent', 'incongruent'))
  
cons_tab_long = cons_tab_long %>% 
  filter(pred_type %in% c('true_acc', 'pred_full_acc')) %>% 
  filter(dec_type!='fcd')
```

```{r}
cons_tab_long  %>% 
  ggplot(aes(x=dec_type, y=value, group=interaction(dec_type,pred_type), color=pred_type, fill = true_bin), 
         position=position_dodge(.9)
         ) +
  geom_boxplot(aes(size=as.factor(true_bin))) +
#  facet_wrap(~judge_type)+
  #theme_classic() +
  #facet_wrap(~con) +
#  scale_size_manual(values=c(1,2))+
  scale_size_manual(values=c(1,1.5))+
  scale_fill_manual(values=c('white','lightgreen')) +
 scale_color_manual(values=rev(c(ccb, cfull, cvu,ctrue)))+
  ylab(label = 'Consistency') +
 # xlab(label = 'judgment') +
  theme_classic(base_size = 32)+
  theme(
    strip.background = element_blank(),
    legend.position = 'none',
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
  ) 
```
### Judgment consistency

```{r}


jcons_tab = x %>% filter(!is.na(judgment), !is.na(judge_type)) %>%
  group_by(subjID, dec_type, judge_type) %>%
  summarise(true = mean(jcons_real, na.rm=T),
            pred_full = mean(jcons_pred_full, na.rm=T),
            pred_cb = mean(jcons_pred_cb, na.rm=T),
            pred_vu = mean(jcons_pred_vu, na.rm=T)) 
jcons_tab %>% 
  group_by(dec_type, judge_type) %>% 
  summarise(real_jcons = mean(true),
            pred_jcons_full = mean(pred_full),
            pred_jcons_cb = mean(pred_cb),
            pred_jcons_vu = mean(pred_vu))
```

```{r}
jcons_tab_long = 
  jcons_tab %>% 
  pivot_longer(cols = c(4:7), 
               values_to = 'cons_value',
               names_to = 'cons_type') 
jcons_tab_long$true_bin=ifelse(jcons_tab_long$cons_type=='true' ,T, F) 

jcons_tab_long = jcons_tab_long %>% filter(cons_type %in% c('true', 'pred_full'))
jcons_tab_long$cons_type = as.factor(jcons_tab_long$cons_type)
#add this weird interaction variable for plotting (judge x con_type)
jcons_tab_long$jxc = interaction(jcons_tab_long$judge_type,jcons_tab_long$cons_type)
a = levels(jcons_tab_long$jxc) #original levels
jcons_tab_long$jxc = factor(jcons_tab_long$jxc ,
                            levels= c(a[3:4] ,a[1:2]))

```

```{r}
pos=position_dodge(.9)
jcons_tab_long  %>% 
  #filter(dec_type==-1) %>% 
  filter(!is.na(dec_type)) %>%
  ggplot(aes(x=judge_type, y=cons_value, group=jxc, color=as_factor(judge_type),linetype=as_factor(cons_type)), 
         position=pos) +
  geom_violin(aes(alpha=cons_type, fill=judge_type),position=pos) +
  geom_boxplot(aes(size=cons_type),position=pos) +
  #facet_wrap(~judge_type)+
  theme_classic() +
 # facet_wrap(~con) +
  scale_alpha_manual(values=c(.5,.8))+
  scale_size_manual(values=c(0.75,1.25))+
  scale_linetype_manual(values=c('dashed', 'solid')) +
  scale_fill_manual(values=c( red1,  blue1)) +
 scale_color_manual(values=c( red1,  blue1))+
  ylab(label = '') +
  xlab(label = '') +
  theme_classic(base_size = 46)+
  theme(
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    legend.position = 'none'
  ) 
```

## Quantitative predictions 

### Choice correlations

```{r}
pred_choice = x %>%
  filter(choice>0) %>%
  filter(con!='fcd') %>% #EXP3: forced choices excluded 
  group_by(subjID, dec_type) %>% summarise(dcor = cor(dpred_full, choice,'complete.obs'))

pred_choice %>% 
  group_by(subjID, dec_type) %>%
  summarise(cor = mean(dcor)) %>% 
  group_by(dec_type) %>% 
  summarise(cor_mean = mean(cor),
            cor_sd = sd(cor))
```
```{r}
pred_choice %>%
  ggplot() +  #aes(x=as.factor(dec_type))
  geom_violin(aes(x=as.factor(dec_type), y=dcor, color=as_factor(dec_type), fill=as_factor(dec_type))) +
  geom_boxplot(aes(x=as.factor(dec_type),y=dcor, color=as_factor(dec_type)), width=.4, size=2) +
  geom_jitter(aes(x=as.factor(dec_type),color=as_factor(dec_type), y=dcor), width=.2, alpha=.3) +
  scale_color_manual(values = rev(c('#235FB9','#D54646'))) +
  scale_fill_manual(values = alpha(rev(c(main_cols[c(4,2)])),.5)) +
  ylim(c(.25,1)) +
  theme(axis.text.y=element_text(size=36))+
  theme(axis.text.x=element_blank())
```


### Judgment correlations
```{r}
pred_judg = x %>%
  filter(judge_type>-1) %>%
  group_by(subjID, judge_type) %>% summarise(jcor = cor(jpred_full, judgment, 'complete.obs'))
```


```{r}
pred_judg %>% 
  group_by(subjID, judge_type) %>%
  summarise(cor = mean(jcor, na.rm=T)) %>% 
  group_by(judge_type) %>% 
  summarise(cor_mean = mean(cor),
            cor_sd = sd(cor))
```
```{r}
pred_judg %>%
  ggplot() +
  geom_violin(aes(x=as.factor(judge_type), y=jcor, color=as_factor(judge_type), fill=as_factor(judge_type))) +
  geom_boxplot(aes(x=as.factor(judge_type),y=jcor, color=as_factor(judge_type)), width=.4, size=2) +
  geom_jitter(aes(x=as.factor(judge_type),color=as_factor(judge_type), y=jcor), width=.2, alpha=.3) +
  scale_color_manual(values = rev(c('#235FB9','#D54646'))) +
  scale_fill_manual(values = alpha(rev(c(main_cols[c(4,2)])),.5)) +
  ylim(c(.25,1)) +
  theme(axis.text.y=element_text(size=36))+
  theme(axis.text.x=element_blank())
```


## Parameters


```{r}
exp_num=3
file_path = paste0(fitted_mods_path,'/params/paramsE',exp_num, '.csv')
pars = read_csv(file_path)
```
we are looking at 3 parameter groups: 
- learning, containing 4 update parameters(alphas, 'V' is pref, and 'P' is size; 'con' is congruent (within-domain) and 'inc' is incongruent (cross-domain))
- cb: consistency bias, containing 3 parameters: intercept jCB, effect of domain jCBdom and effect of congruency jCBcon
- value: related to value influence on judgment: effect of difference jDIF and effect of sum jDxS.


```{r}
pars
```
We can plot given parameters, e.g. learning within domain:
```{r}
pars %>%
  filter(par_class=='learning') %>% 
  filter(congruency=='con') %>%
  ggplot(aes(x=parameter, y=value, fill=parameter))+
  ggdist::stat_eye(alpha=.6) +
 # scale_x_discrete(labels = c(TeX('$u^{pref}_{con}$'), TeX("$u^{size}_{con}$"))) +
  scale_fill_manual(values = c(red1,blue1)) +
  theme(text=element_text(size=20))
```
...cross-domain...
```{r}
pars %>%
  filter(par_class=='learning') %>% 
  filter(congruency=='inc') %>%
  ggplot(aes(x=parameter, y=value, fill=parameter))+
  ggdist::stat_eye(alpha=.6) +
 # scale_x_discrete(labels = c(TeX('$u^{pref}_{con}$'), TeX("$u^{size}_{con}$"))) +
  scale_fill_manual(values = c(red1,blue1)) +
  theme(text=element_text(size=20))
```

or others
```{r}
pars %>%
  filter(par_class=='cb') %>% 
  ggplot(aes(x=parameter, y=value, fill=parameter))+
  ggdist::stat_eye(alpha=.6) +
 # scale_x_discrete(labels = c(TeX('$u^{pref}_{con}$'), TeX("$u^{size}_{con}$"))) +
 # scale_fill_manual(values = c(red1,blue1)) +
  theme(text=element_text(size=20))
```

